# Treatment Status {#sec-treat}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

This chapter shows how to visualize the treatment conditions and missing values in a panel dataset using `type = "treat"` and `type = "missing"`. The treatment indicator may be dichotomous, multi-level, or continuous.

We first load the package and its built-in datasets.

```{r}
library(panelView)
data(panelView)
ls()
```

## Binary treatment

Using the `turnout` dataset (a balanced panel), we visualize the treatment status of Election Day Registration (EDR) across US states ([Xu 2017](http://dx.doi.org/10.1017/pan.2016.2)). The first variable on the right-hand side of the formula is the treatment indicator. The `index` option specifies the unit and time variables.

```{r treat1-1}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State")
```

Use `by.timing = TRUE` to sort units by the timing of first treatment. The `background` option sets the panel background color; `cex.*` options control font sizes.

```{r treat1-2, fig.height=10}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", by.timing = TRUE,
          legend.labs = c("No EDR", "EDR"), background = "white",
          cex.main = 20, cex.axis = 8, cex.lab = 12, cex.legend = 12)
```

For staggered adoption, `pre.post = TRUE` distinguishes pre- and post-treatment periods for treated units.

```{r treat1-3, fig.height=10}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", pre.post = TRUE)
```

Customise legend labels by passing a character vector to `legend.labs`. Its length must match the number of distinct treatment statuses shown.

```{r treat1-4, fig.height=8}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", by.timing = TRUE,
          pre.post = TRUE,
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

Control axis label display with `axis.lab`: `"both"` (default), `"time"`, `"unit"`, or `"off"`.

```{r treat1-5, fig.height=8}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          main = "EDR Reform", axis.lab = "time")
```

Instead of a formula, supply the treatment variable name directly via `D =`.

```{r treat1-6, eval=FALSE}
panelview(D = "policy_edr", data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", pre.post = TRUE,
          main = "EDR Reform", axis.lab = "time")
```

Override brick colors with the `color` option. When `pre.post = FALSE`, colors follow the order `c("control", "treated")`; when `pre.post = TRUE`, the order is `c("treated-pre", "treated-post", "control")`.

```{r treat1-7, fig.height=8}
library(dplyr)
turnout |>
  panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
            color = c("#B0C4DE", "white"), by.timing = TRUE,
            index = c("abb", "year"), xlab = "Year", ylab = "State")
```

When the time variable has gaps, use `leave.gap = TRUE` to display them as white bars. Without it, **panelView** skips the gaps and issues a warning.

```{r treat1-8, fig.height=8}
turnout2 <- turnout[!(turnout$year %in% c("1924", "1928", "1940")), ]
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout2, index = c("abb", "year"),
          type = "treat", leave.gap = TRUE)
```

## Switch-on/switch-off treatment

The `capacity` dataset has treatment that can turn on and off (democratic reversals). In this setting, `pre.post` is not meaningful.

```{r treat2-1, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity",
          axis.lab.gap = c(2, 10))
```

Sorting by `by.timing = TRUE` and rotating x-axis labels with `axis.lab.angle = 90` often improves readability.

```{r treat2-2, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity: Treatment Status",
          axis.lab.angle = 90, by.timing = TRUE, axis.lab = "time")
```

## Collapsing units by treatment history

When the number of units is large, `collapse.history = TRUE` collapses units that share the same treatment history into a single row. The y-axis shows the group size.

```{r treat2-3, fig.height=4}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), axis.lab.angle = 90,
          by.timing = TRUE, collapse.history = TRUE)
```

```{r treat2-4, fig.height=8}
panelview(Capacity ~ demo + lnpop + lngdp, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = c(2, 0),
          axis.lab.angle = 90, collapse.history = TRUE)
```

## Plotting a subset of units

Use `show.id` to select units by their rank (alphabetical order), or `id` to select by their original identifiers.

```{r treat3-1, fig.height=5}
panelview(Capacity ~ demo + lnpop + lngdp, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = c(2, 0),
          main = "Democracy and State Capacity", show.id = 1:25)
```

```{r treat3-3, fig.height=5}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 0),
          main = "Democracy and State Capacity",
          id = unique(capacity$ccode)[26:50], by.timing = TRUE)
```

## Missingness only

Set `type = "missing"` to plot only the pattern of missing values, ignoring treatment assignment.

```{r treat3-4, fig.height=10}
capacity |>
  panelview(Capacity ~ 1,
            index = c("ccode", "year"),
            axis.lab = "off", type = "missing")
```

Supply a variable name instead of a formula via `Y =`.

```{r treat3-6, eval=FALSE}
capacity |>
  panelview(Y = "Capacity",
            index = c("ccode", "year"),
            axis.lab.gap = c(2, 10),
            type = "missing")
```

With `leave.gap = TRUE`, time gaps are visible even when the data has no explicit gap rows.

```{r treat3-7, fig.height=10}
capacity2 <- capacity
capacity2$demo[capacity2$year %in% c("1960", "1980")] <- NA
capacity2$lngdp[capacity2$year == "1990"] <- NA
panelview(1 ~ demo + lngdp, data = capacity2, index = c("ccode", "year"),
          type = "missing", leave.gap = TRUE, axis.lab.gap = c(2, 10))
```

## Multi-level treatment

**panelView** supports treatment variables with more than 2 levels.

```{r treat4-1, fig.height=10}
demo2 <- rep(0, nrow(capacity))
demo2[capacity$polity2 < -0.5] <- -1
demo2[capacity$polity2 >  0.5] <-  1
capacity$demo2 <- demo2

panelview(Capacity ~ demo2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10),
          main = "Regime Type")
```

Assign custom colors and legend labels.

```{r treat4-2, fig.height=10}
library(RColorBrewer)
mycol <- brewer.pal(3, "Set1")[c(1, 3, 2)]
panelview(Capacity ~ demo2, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = 2,
          main = "Regime Type", axis.lab = "time",
          color = mycol,
          legend.labs = c("Autocracy", "Hybrid", "Democracy"))
```

## Continuous treatment

When the treatment has more than 5 unique values, **panelView** treats it as continuous and renders a gradient heatmap. Use `gridOff = TRUE` to remove grid lines.

```{r treat4-4, fig.height=10}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10), main = "Regime Type",
          gridOff = TRUE)
```

Customise the gradient palette via `color`.

```{r treat4-5, fig.height=10}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10), main = "Regime Type",
          color = c("yellow", "orange", "red", "purple", "brown"),
          background = "white")
```
