# Treatment Status {#sec-treat}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)
```

This chapter shows how to visualize the treatment conditions and missing values in a panel dataset using `type = "treat"` and `type = "missing"`. The treatment indicator may be dichotomous, multi-level, or continuous.

We first load the package and its built-in datasets.

```{r}
library(panelView)
data(panelView)
ls()
```

## Binary treatment

Using the `turnout` dataset (a balanced panel), we visualize the treatment status of Election Day Registration (EDR) across US states ([Xu 2017](http://dx.doi.org/10.1017/pan.2016.2)). The first variable on the right-hand side of the formula is the treatment indicator. The `index` option specifies the unit and time variables.

```{r treat1-1}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State")
```

Use `by.timing = TRUE` to sort units by the timing of first treatment.

```{r treat1-2, fig.height=10}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", by.timing = TRUE,
          legend.labs = c("No EDR", "EDR"))
```

For staggered adoption, `pre.post = TRUE` distinguishes pre- and post-treatment periods for treated units.

```{r treat1-3, fig.height=10}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", pre.post = TRUE)
```

Customise legend labels by passing a character vector to `legend.labs`. Its length must match the number of distinct treatment statuses shown.

```{r treat1-4, fig.height=8}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", by.timing = TRUE,
          pre.post = TRUE,
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

Control axis label display with `axis.lab`: `"both"` (default), `"time"`, `"unit"`, or `"off"`.

```{r treat1-5, fig.height=8}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          main = "EDR Reform", axis.lab = "time")
```

Instead of a formula, supply the treatment variable name directly via `D =`.

```{r treat1-6, eval=FALSE}
panelview(D = "policy_edr", data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", pre.post = TRUE,
          main = "EDR Reform", axis.lab = "time")
```

Override brick colors with the `color` option. When `pre.post = FALSE`, colors follow the order `c("control", "treated")`; when `pre.post = TRUE`, the order is `c("treated-pre", "treated-post", "control")`.

```{r treat1-7, fig.height=8}
library(dplyr)
turnout |>
  panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
            color = c("#B0C4DE", "white"), by.timing = TRUE,
            index = c("abb", "year"), xlab = "Year", ylab = "State")
```

When the time variable has gaps, use `leave.gap = TRUE` to display them as white bars. Without it, **panelView** skips the gaps and issues a warning.

```{r treat1-8, fig.height=8}
turnout2 <- turnout[!(turnout$year %in% c("1924", "1928", "1940")), ]
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout2, index = c("abb", "year"),
          type = "treat", leave.gap = TRUE)
```

## Switch-on/switch-off treatment

The `capacity` dataset has treatment that can turn on and off (democratic reversals). In this setting, `pre.post` is not meaningful.

```{r treat2-1, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity",
          axis.lab.gap = c(2, 10))
```

Sorting by `by.timing = TRUE` and rotating x-axis labels with `axis.lab.angle = 90` often improves readability.

```{r treat2-2, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity: Treatment Status",
          axis.lab.angle = 90, by.timing = TRUE, axis.lab = "time")
```

## Themes

The `theme` argument selects the color recipe. The default (`"default"`)
uses panelView's classic blue progression. Setting `theme = "red"`
activates a high-contrast publication-paper recipe: control units fade to
a light gray and treated post-period cells pick up a brick-red accent.

```{r treat2-theme-red, fig.height=8}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          xlab = "Year", ylab = "State", by.timing = TRUE,
          pre.post = TRUE, theme = "red",
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

The `red` theme also handles panels with missing observations. On
`capacity` (treatment turns on and off; some unit-year cells are
unobserved), the heatmap reads as three layers --- gray untreated cells,
brick-red treated cells, and white missing cells --- with all three shown
in the legend.

```{r treat2-theme-red-missing, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity",
          axis.lab.gap = c(2, 10), theme = "red")
```

If the dusty pink for "Under Control" feels too warm, swap in a neutral
light grey via the named `color` argument so the dark-red treatment
cells become the only chromatic element. All other palette slots
inherit the red theme's defaults:

```{r treat2-theme-red-grey, fig.height=10}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Democracy and State Capacity",
          axis.lab.gap = c(2, 10), theme = "red",
          color = c(control = "grey85"))
```

## Customizing with ggplot2 syntax

`panelview()` returns a `ggplot` object for single-panel plots, so any
`ggplot2` layer can be added with the usual `+`. In v1.3.1 the default
title is left-aligned and plain; to recover a centered, bold title, add a
`theme(plot.title = ...)` layer:

```{r treat2-ggplot-center-title, fig.height=8}
library(ggplot2)
p <- panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
               data = turnout, index = c("abb", "year"),
               main = "Election-Day Registration Adoption",
               by.timing = TRUE, theme = "red")
p + theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The same pattern works for other theme elements (axis text, legend
position, margins, etc.). Multi-panel layouts (`by.group`,
`by.group.side`, `by.cohort`, `by.unit`) are an exception: those return a
`gtable` produced by `grid.arrange()` and do not accept `+` composition.

## Collapsing units by treatment history

When the number of units is large, `collapse.history = TRUE` collapses units that share the same treatment history into a single row. The y-axis shows the group size.

```{r treat2-3, fig.height=4}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), axis.lab.angle = 90,
          by.timing = TRUE, collapse.history = TRUE)
```

```{r treat2-4, fig.height=8}
panelview(Capacity ~ demo + lnpop + lngdp, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = c(2, 0),
          axis.lab.angle = 90, collapse.history = TRUE)
```

## Plotting a subset of units

Use `show.id` to select units by their rank (alphabetical order), or `id` to select by their original identifiers.

```{r treat3-1, fig.height=5}
panelview(Capacity ~ demo + lnpop + lngdp, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = c(2, 0),
          main = "Democracy and State Capacity", show.id = 1:25)
```

```{r treat3-3, fig.height=5}
panelview(Capacity ~ demo + lnpop + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 0),
          main = "Democracy and State Capacity",
          id = unique(capacity$ccode)[26:50], by.timing = TRUE)
```

## Missingness only

Set `type = "missing"` to plot only the pattern of missing values, ignoring treatment assignment.

```{r treat3-4, fig.height=10}
capacity |>
  panelview(Capacity ~ 1,
            index = c("ccode", "year"),
            axis.lab = "off", type = "missing")
```

Supply a variable name instead of a formula via `Y =`.

```{r treat3-6, eval=FALSE}
capacity |>
  panelview(Y = "Capacity",
            index = c("ccode", "year"),
            axis.lab.gap = c(2, 10),
            type = "missing")
```

With `leave.gap = TRUE`, time gaps are visible even when the data has no explicit gap rows.

```{r treat3-7, fig.height=10}
capacity2 <- capacity
capacity2$demo[capacity2$year %in% c("1960", "1980")] <- NA
capacity2$lngdp[capacity2$year == "1990"] <- NA
panelview(1 ~ demo + lngdp, data = capacity2, index = c("ccode", "year"),
          type = "missing", leave.gap = TRUE, axis.lab.gap = c(2, 10))
```

## Multi-level treatment

**panelView** supports treatment variables with more than 2 levels.

```{r treat4-1, fig.height=10}
demo2 <- rep(0, nrow(capacity))
demo2[capacity$polity2 < -0.5] <- -1
demo2[capacity$polity2 >  0.5] <-  1
capacity$demo2 <- demo2

panelview(Capacity ~ demo2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10),
          main = "Regime Type")
```

Assign custom colors and legend labels.

```{r treat4-2, fig.height=10}
library(RColorBrewer)
mycol <- brewer.pal(3, "Set1")[c(1, 3, 2)]
panelview(Capacity ~ demo2, data = capacity,
          index = c("ccode", "year"), axis.lab.gap = 2,
          main = "Regime Type", axis.lab = "time",
          color = mycol,
          legend.labs = c("Autocracy", "Hybrid", "Democracy"))
```

## Continuous treatment

When the treatment has more than 5 unique values, **panelView** treats it as continuous and renders a gradient heatmap. Use `gridOff = TRUE` to remove grid lines.

```{r treat4-4, fig.height=10}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10), main = "Regime Type",
          gridOff = TRUE)
```

Customise the gradient palette via `color`.

```{r treat4-5, fig.height=10}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          axis.lab.gap = c(2, 10), main = "Regime Type",
          color = c("yellow", "orange", "red", "purple", "brown"),
          background = "white")
```

## Overlaying the estimation sample {#sec-sample}

For the treatment plot (`type = "treat"`), **panelView** can shade
the cells the estimator actually used in grey. This shows what
fraction of the panel a given method relies on, and which units or
periods are dropped. The feature works for the imputation estimators
**fect**, **gsynth**, and **tjbal**, and for the matching estimator
**PanelMatch**.

The example data in this section is the `dem` panel from
**PanelMatch**: 175 countries from 1960 to 2010, with a binary
democracy indicator and per-capita income (`y`) as outcome. The data
are already in long format. `PanelMatch::PanelData()` attaches
treatment, outcome, unit-id, and time-id metadata to the data.frame
so that subsequent `PanelMatch()` calls can resolve those columns
without restating them; it does not reshape the data. `fect()` and
`panelview()` treat the result, `pd`, as an ordinary data.frame.

```{r treat-data-setup, message=FALSE, warning=FALSE}
library(PanelMatch); library(fect); library(panelView)
data(dem)
pd <- PanelMatch::PanelData(panel.data = dem, unit.id = "wbcode2",
                            time.id = "year", treatment = "dem",
                            outcome = "y")
```

### Baseline: pre-fit treatment plot

Before fitting any estimator, the standard treatment plot shows the
status of every cell:

```{r treat-prefit, fig.height=10}
panelview(y ~ dem, data = pd, index = c("wbcode2", "year"),
          type = "treat", by.timing = TRUE, axis.lab = "off")
```

By default, dropped controls and dropped treated units share a single
neutral grey. The control vs. treated distinction is already shown by
the colors of used cells, so a single unused color keeps the main
contrast in the figure between what the estimator used and what it
did not. A two-color split for unused cells is available as an option;
the PanelMatch section below shows when it is useful.

### PanelMatch

Imai, Kim, and Wang's `PanelMatch` matches each treated event to a
set of control units on the basis of pre-treatment history. Its
output is a list of matched sets rather than a fitted panel of
counterfactuals. The helper `panelmatch_to_sample()` turns those
matched sets into a $T \times N$ logical matrix that `panelview()`
can shade. It reads only the public fields of `PanelMatch`
(`pm$att`, `pm$atc`, and the `unit.id`, `time.id`, `lag`, `lead`
attributes), so no changes to `PanelMatch` are needed.

Run `PanelMatch` on the same `pd`:

```{r treat-pm-setup, message=FALSE}
pm <- PanelMatch(panel.data = pd, lag = 4,
                 refinement.method = "mahalanobis",
                 covs.formula = ~ I(lag(tradewb, 1:4)),
                 size.match = 5,
                 qoi = "att", lead = 0:3, match.missing = TRUE)
```

#### Pass the matched-set object directly

```{r treat-pm-direct, fig.height=10}
panelview(data = pd, formula = y ~ dem,
          index = c("wbcode2", "year"), type = "treat",
          by.timing = TRUE,
          sample = pm,
          main = "PanelMatch mask: full window [t-4, t+3]",
          axis.lab = "off")
```

`panelview()` recognises a `PanelMatch` object and calls
`panelmatch_to_sample()` automatically. The helper marks the window
$[t-\text{lag},\,t+\max(\text{lead})]$ around each treated event $t$:
the lag part covers periods used for matching, the lead part covers
periods on which outcomes are evaluated. Within each event the
light-blue pre-period and dark-blue post-onset cells therefore appear
as one continuous bright band.

Bright cells appear in at least one matched-set window. Grey cells
fall outside every window. The grey area has several distinct
sources: never-treated countries that never served as a matched
control under this refinement; treated countries whose event window
extends past the start or end of the panel; and periods of in-sample
units that sit outside any nearby event.

#### Split unused into control and treated

To separate dropped controls from dropped treated units in the
figure, pass different colors to `unused.control` and
`unused.treated`:

```{r treat-pm-split, fig.height=10}
panelview(data = pd, formula = y ~ dem,
          index = c("wbcode2", "year"), type = "treat",
          by.timing = TRUE, sample = pm,
          color = c(unused.control = "#D9D9D9",
                    unused.treated = "#7A7A7A"),
          axis.lab = "off")
```

The legend now shows "Not used: Under Control" and "Not used: Under
Treatment" as two entries instead of one. Setting both colors to the
same value collapses them back to a single "Not used" entry. Under
PanelMatch both kinds are visible: the light-grey rows are
never-democratised countries that never served as a matched control
under this refinement, and the darker-grey rows are always-democratised
countries that had no valid event to match.

#### Customise the full palette

All five used-cell colors are controlled through the same `color =`
argument, which accepts a named vector. Any name not supplied keeps
its theme default:

```{r treat-pm-color, fig.height=10}
panelview(data = pd, formula = y ~ dem,
          index = c("wbcode2", "year"), type = "treat",
          by.timing = TRUE, sample = pm,
          color = c(control        = "#A8C686",
                    treated        = "#D87C2A",
                    missing        = "white",
                    unused.control = "#F0E0CC",
                    unused.treated = "#F0E0CC"),
          axis.lab = "off")
```

Allowed names are `control`, `treated`, `treated.pre` (three-state
binary only), `missing`, `unused.control`, and `unused.treated`. An
unnamed positional vector is also accepted, preserving the older
`color = c("#color1", "#color2", ...)` form.

To change how matched sets are turned into the mask, call
`panelmatch_to_sample(pm, pd)` directly. Useful arguments are
`include` (limit the mask to treated units or matched controls only),
`weight.threshold` (set the cutoff for positive weights when the
refinement returns continuous weights), and `qoi = "atc"` (use the
ATC matched sets when `pm` carries both).

#### A wider matching spec

Doubling the matching window to `lag = 8` and extending the outcome
window to `lead = 0:6` widens each event's bright band from 8 cells
to 15. It also drops events too close to either edge of the panel:
events before 1968 have no $[t-8, t-1]$ history, and events after 2004
have no $[t, t+6]$ leads.

```{r treat-pm-wide, fig.height=10, message=FALSE}
pm_wide <- PanelMatch(panel.data = pd, lag = 8,
                      refinement.method = "mahalanobis",
                      covs.formula = ~ I(lag(tradewb, 1:8)),
                      size.match = 5,
                      qoi = "att", lead = 0:6, match.missing = TRUE)
panelview(data = pd, formula = y ~ dem,
          index = c("wbcode2", "year"), type = "treat",
          by.timing = TRUE, sample = pm_wide,
          main = "PanelMatch mask: lag = 8, lead = 0:6",
          axis.lab = "off")
```

### Imputation estimators (fect / gsynth)

```{r treat-fit-setup, message=FALSE, warning=FALSE}
fit <- fect(y ~ dem, data = pd, index = c("wbcode2", "year"),
            method = "ife", r = 1, se = FALSE, CV = FALSE,
            min.T0 = 2, na.rm = TRUE)
```

#### Pass the fitted object directly

```{r treat-fit-1, fig.height=10}
panelview(fit, type = "treat", by.timing = TRUE,
          axis.lab = "off", display.all = TRUE)
```

No `data`, no `formula`, no `index` is required: `panelview()` reads
them from the fit object. This is the simplest call because the
figure shows exactly the panel the estimator used; the data shown in
the plot cannot differ from the data used in estimation.

Reading the figure from top to bottom: a grey "Not used" band of
always-treated countries (democratic since 1960, with no control
periods for fect to use) and a smaller number of never-treated
countries whose outcome is too sparse to fit a counterfactual; then
a dark-blue staircase of in-sample treated units ordered by year of
transition, with light-blue cells marking their pre-transition years;
and finally a light-blue block of never-democratised in-sample
controls. White cells are originally missing outcomes.

#### Red theme

`theme = "red"` switches to a higher-contrast palette intended for
print:

```{r treat-fit-red, fig.height=10}
panelview(fit, type = "treat", by.timing = TRUE,
          theme = "red", axis.lab = "off", display.all = TRUE)
```

The red theme uses dusty pink (`#E8B5BC`) for "Under Control" and
dark red (`#B83A4B`) for "Under Treatment". Cells in either treatment
state share a warm hue; grey "Not used" cells sit visually apart.
White cells indicate original missingness.

To put the in-sample band at the top of the figure instead of the
dropped band, set `sample.sort = TRUE`. Combined with `by.timing`,
the four options cover the common row orderings:

| `by.timing` | `sample.sort` | Row order |
|---|---|---|
| `FALSE` | `FALSE` | alphabetical / natural order |
| `TRUE`  | `FALSE` | sorted by treatment timing only (default for fits) |
| `FALSE` | `TRUE`  | in-sample first, out-of-sample below |
| `TRUE`  | `TRUE`  | in-sample first, by treatment timing within |

#### A stricter spec drops more units

The shading is also useful for checking how sensitive the in-sample
set is to specification choices. Raising `min.T0` from 2 to 10
requires every kept unit to have at least 10 untreated periods; about
14 more countries fail this requirement and the "Not used" band
widens accordingly:

```{r treat-fit-strict, fig.height=10, message=FALSE, warning=FALSE}
fit_strict <- fect(y ~ dem, data = pd, index = c("wbcode2", "year"),
                   method = "ife", r = 1, se = FALSE, CV = FALSE,
                   min.T0 = 10, na.rm = TRUE)
panelview(fit_strict, type = "treat", by.timing = TRUE,
          axis.lab = "off", display.all = TRUE)
```

The in-sample share of cells drops from about 56% to about 51%.
The "Not used" band at the top thickens compared with `fit` above.
Originally missing cells remain white.

#### Pass `sample = fit` alongside explicit data + formula

The original `(data, formula, index)` form still works, and
`sample =` accepts the fit object directly. The figure matches the
fit-as-input view above. This form is useful when the original `data`
object is not loaded in the current session --- for example, after
reading a saved `.rds` fit:

```{r treat-sample-explicit, eval=FALSE}
panelview(y ~ dem, data = pd, index = c("wbcode2", "year"),
          type = "treat", by.timing = TRUE,
          sample = fit,
          axis.lab = "off")
```

`sample =` accepts any `fect`, `gsynth`, or `tjbal` fit that
exposes a `$sample` field; the same call works across the three
packages. A logical $T \times N$ matrix can also be passed directly,
which makes the feature available to custom estimators.

**Alignment by name.** The `sample` matrix and the panel data must
agree on which units and times appear. `panelview()` follows two
asymmetric rules:

- *Data has units or periods the sample does not.* The extra cells
  are filled with `FALSE`. This is useful when the estimator was fit
  on a subset of a larger panel; the extra units appear in the
  "Not used" band.
- *Sample has units or periods the data does not.* `panelview()`
  stops with an error that lists the mismatched cells. This usually
  means the sample is stale or came from a different fit.

If `sample` has `rownames` and `colnames`, those are used to align by
name. A bare matrix without dimnames is aligned by position.
