# Outcome Trajectories {#sec-outcome}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
library(panelView)
data(panelView)
```

This chapter shows how to plot raw outcome variables as time series, colored by treatment status, using `type = "outcome"`. The syntax is the same as for `type = "treat"` with `type = "outcome"` added.

## Basic outcome plot

```{r outcome1-1}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Turn off pre/post coloring with `pre.post = FALSE`.

```{r outcome1-1a}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), pre.post = FALSE,
          type = "outcome", main = "EDR Reform and Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Customise the legend, or suppress it entirely with `legendOff = TRUE`.

```{r outcome1-2}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

Switch off the black-and-white theme with `theme.bw = FALSE`.

```{r outcome1-3}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout", theme.bw = FALSE)
```

Customise line colors and legend labels simultaneously.

```{r outcome1-4}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          color = c("lightblue", "blue", "#99999950"),
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

## Plotting a subset of units

Use `id` to plot only specific units.

```{r outcome1-5}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout (AL, AR, CT)",
          id = c("AL", "AR", "CT"))
```

## Splitting by treatment-status group

`by.group = TRUE` separates units into groups based on whether their treatment status ever changed (e.g., always treated, always control, switchers). `cex.main` and `cex.main.sub` control title and subtitle font sizes.

```{r outcome1-6}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.group = TRUE, cex.main = 20, cex.main.sub = 15)
```

Use `by.group.side = TRUE` to arrange the subfigures in a row rather than a column.

```{r outcome1-7}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.group.side = TRUE, cex.main = 20, cex.main.sub = 15)
```

## Cohort-averaged trajectories

`by.cohort = TRUE` collapses units with identical treatment histories into cohort-average trajectories. This helps diagnose treatment-effect heterogeneity that may bias TWFE estimates.

```{r outcome1-4a}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.cohort = TRUE)
```

A more detailed cohort plot with custom legend labels and y-axis limits.

```{r outcome4-1}
panelview(turnout ~ policy_edr + policy_motor,
          data = turnout, index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout", by.cohort = TRUE,
          ylim = c(40, 80),
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

## Ignoring treatment status

Since v1.0.3, omitting the treatment indicator (using `Y ~ 1`) lets **panelView** visualize any variable in a panel dataset without reference to treatment.

```{r outcome2-1}
panelview(turnout ~ 1, data = turnout, index = c("abb", "year"),
          type = "outcome", main = "Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Equivalently, supply the variable name via `Y =`.

```{r outcome2-2, eval=FALSE}
panelview(Y = "turnout", data = turnout,
          index = c("abb", "year"), type = "outcome",
          main = "Turnout", ylim = c(0, 100),
          xlab = "Year", ylab = "Turnout")
```

## Discrete outcomes

Set `outcome.type = "discrete"` when the outcome takes a small number of integer values.

```{r outcome3-1, fig.height=5}
panelview(Y ~ D, data = simdata, index = c("id", "time"),
          outcome.type = "discrete",
          type = "outcome", xlim = c(8, 15))
```

Split by treatment-status group.

```{r outcome3-2, fig.height=5}
panelview(Y ~ D, data = simdata, index = c("id", "time"),
          by.group = TRUE, outcome.type = "discrete",
          type = "outcome", xlim = c(8, 15))
```

## Multi-level or continuous treatment

When the treatment variable has more than 5 unique values (e.g., `polity2`), treatment status is not shown on the outcome plot.

```{r outcome3-3, fig.height=5}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Measuring State Capacity",
          type = "outcome", legendOff = TRUE)
```
