# Outcome Trajectories {#sec-outcome}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
library(panelView)
data(panelView)
```

This chapter shows how to plot raw outcome variables as time series, colored by treatment status, using `type = "outcome"`. The syntax is the same as for `type = "treat"` with `type = "outcome"` added.

## Basic outcome plot

```{r outcome1-1}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Turn off pre/post coloring with `pre.post = FALSE`.

```{r outcome1-1a}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), pre.post = FALSE,
          type = "outcome", main = "EDR Reform and Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Customise the legend, or suppress it entirely with `legendOff = TRUE`.

```{r outcome1-2}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

Customise line colors and legend labels simultaneously.

```{r outcome1-4}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          color = c("lightblue", "blue", "grey75"),
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

## Themes and overlays

Activate the high-contrast `red` theme for a publication-paper look. Under
`theme = "red"`, controls fade to a light gray, treated trajectories pick
up a brick-red accent, and the treatment-onset marker becomes a solid
black dashed line.

```{r outcome1-3}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout", theme = "red")
```

Overlay heavy group-mean trajectories and a 10--90% quantile ribbon with
`group.mean.overlay = TRUE`. Useful when there are many units per group
and you want the audience to see both individual variation and the group
average at a glance. (Currently implemented for the main DID continuous-
outcome path; combine with `theme = "red"` for the cleanest read.)

```{r outcome1-3b}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout",
          theme = "red", group.mean.overlay = TRUE)
```

## Plotting a subset of units

Use `id` to plot only specific units.

```{r outcome1-5}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout (AL, AR, CT)",
          id = c("AL", "AR", "CT"))
```

## Splitting by treatment-status group

`by.group = TRUE` separates units into groups based on whether their treatment status ever changed (e.g., always treated, always control, switchers).

```{r outcome1-6}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.group = TRUE)
```

Use `by.group.side = TRUE` to arrange the subfigures in a row rather than a column.

```{r outcome1-7}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.group.side = TRUE)
```

## Cohort-averaged trajectories

`by.cohort = TRUE` collapses units with identical treatment histories into cohort-average trajectories. This helps diagnose treatment-effect heterogeneity that may bias TWFE estimates.

```{r outcome1-4a}
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor,
          data = turnout, index = c("abb", "year"),
          type = "outcome", main = "EDR Reform and Turnout",
          by.cohort = TRUE)
```

A more detailed cohort plot with custom legend labels and y-axis limits.

```{r outcome4-1}
panelview(turnout ~ policy_edr + policy_motor,
          data = turnout, index = c("abb", "year"), type = "outcome",
          main = "EDR Reform and Turnout", by.cohort = TRUE,
          ylim = c(40, 80),
          legend.labs = c("Control States",
                          "Treated States (before EDR)",
                          "Treated States (after EDR)"))
```

## Ignoring treatment status

Since v1.0.3, omitting the treatment indicator (using `Y ~ 1`) lets **panelView** visualize any variable in a panel dataset without reference to treatment.

```{r outcome2-1}
panelview(turnout ~ 1, data = turnout, index = c("abb", "year"),
          type = "outcome", main = "Turnout",
          ylim = c(0, 100), xlab = "Year", ylab = "Turnout")
```

Equivalently, supply the variable name via `Y =`.

```{r outcome2-2, eval=FALSE}
panelview(Y = "turnout", data = turnout,
          index = c("abb", "year"), type = "outcome",
          main = "Turnout", ylim = c(0, 100),
          xlab = "Year", ylab = "Turnout")
```

## Treatment reversal and missing data {#sec-outcome-reversal}

How **panelView** handles two common discontinuities in the outcome
trajectory:

- **Treatment reversal** (unit goes on then off, or off then on). The
  treated overlay is drawn only at periods the unit is treated, so each
  treated spell is a separate line segment.
- **Missing outcome**. Under the default `leave.gap = FALSE`, rows with
  missing `Y` are dropped and the bracketing observed points are joined
  by a **dotted** segment so the gap is visible. Under `leave.gap = TRUE`
  the panel is expanded and the line breaks cleanly at every NA cell.

Topology classification (reversal vs. staggered) follows the observed
data: if a unit's leading treated rows are missing and the surviving
`D` pattern is monotonic, the unit is plotted in the staggered palette.

Examples below use `fect::sim_base`. Unit 102 is treated at $t=1\!-\!3$,
control at $t=4\!-\!26$, then treated at $t=27\!-\!35$.

### Reversal alone

```{r outcome-reversal, fig.height=4.2}
if (requireNamespace("fect", quietly = TRUE)) {
  data("sim_base", package = "fect")
  panelview(Y ~ D, data = subset(sim_base, id == 102),
            index = c("id", "time"), type = "outcome",
            main = "Unit 102: treated 1-3, control 4-26, treated 27-35")
}
```

### Interior gap: dotted bridge under `leave.gap = FALSE`

Set $Y$ to `NA` at $t=10\!-\!12$ and $t=30\!-\!31$. The control line
shows a dotted bridge between $t=9$ and $t=13$; the treated line shows
one between $t=29$ and $t=32$.

```{r outcome-missing-default, fig.height=4.2}
if (requireNamespace("fect", quietly = TRUE)) {
  d <- subset(sim_base, id == 102)
  d$Y[d$time %in% c(10, 11, 12)] <- NA
  d$Y[d$time %in% c(30, 31)]     <- NA
  panelview(Y ~ D, data = d, index = c("id", "time"),
            type = "outcome",
            main = "leave.gap = FALSE: dotted bridge across missing periods")
}
```

### Interior gap: clean break under `leave.gap = TRUE`

Same data, but `leave.gap = TRUE` keeps the NA rows, so the line breaks
at every gap with no bridge.

```{r outcome-missing-leave-gap, fig.height=4.2}
if (requireNamespace("fect", quietly = TRUE)) {
  d <- subset(sim_base, id == 102)
  d$Y[d$time %in% c(10, 11, 12)] <- NA
  d$Y[d$time %in% c(30, 31)]     <- NA
  panelview(Y ~ D, data = d, index = c("id", "time"),
            type = "outcome", leave.gap = TRUE,
            main = "leave.gap = TRUE: line breaks at gaps, no bridge")
}
```

### Missing at the boundary

Drop unit 102's $Y$ at $t=1\!-\!3$ (the start of its series) and pass
`leave.gap = TRUE` so the row is retained, the x-axis stays at
$t=1\!-\!35$, and the topology classifier still sees the early
treated $D$ values. The line is empty at $t=1\!-\!3$; everything else
is unchanged.

```{r outcome-missing-boundary, fig.height=4.2}
if (requireNamespace("fect", quietly = TRUE)) {
  d <- subset(sim_base, id == 102)
  d$Y[d$time %in% 1:3] <- NA
  panelview(Y ~ D, data = d, index = c("id", "time"),
            type = "outcome", leave.gap = TRUE,
            main = "Drop Y at t=1-3 (leave.gap = TRUE keeps x-axis and classification)")
}
```

Under the default `leave.gap = FALSE`, the same NA injection drops
those rows entirely. The panel for this unit shrinks to $t=4\!-\!35$,
its surviving $D$ pattern becomes monotonic, and the classifier
re-types it as staggered. Use `leave.gap = TRUE` whenever you need
the boundary intact.

### `by.group = TRUE` with all three subpanels

`by.group = TRUE` splits units by their treatment history: always under
control, always under treatment, and treatment status changed. To
exercise all three subpanels we promote one `sim_base` unit to
always-treated.

```{r outcome-bygroup-full, fig.height=8}
if (requireNamespace("fect", quietly = TRUE)) {
  d <- sim_base
  d$D[d$id == 101] <- 1    ## promote unit 101 to always-treated
  panelview(Y ~ D, data = d, index = c("id", "time"),
            type = "outcome", by.group = TRUE,
            main = "by.group = TRUE: always control / always treated / status changed")
}
```

## Discrete outcomes

Set `outcome.type = "discrete"` when the outcome takes a small number of integer values.

```{r outcome3-1, fig.height=5}
panelview(Y ~ D, data = simdata, index = c("id", "time"),
          outcome.type = "discrete",
          type = "outcome", xlim = c(8, 15))
```

Split by treatment-status group.

```{r outcome3-2, fig.height=5}
panelview(Y ~ D, data = simdata, index = c("id", "time"),
          by.group = TRUE, outcome.type = "discrete",
          type = "outcome", xlim = c(8, 15))
```

## Multi-level or continuous treatment

When the treatment variable has more than 5 unique values (e.g., `polity2`), treatment status is not shown on the outcome plot.

```{r outcome3-3, fig.height=5}
panelview(Capacity ~ polity2 + lngdp,
          data = capacity, index = c("ccode", "year"),
          main = "Measuring State Capacity",
          type = "outcome", legendOff = TRUE)
```
