# Welcome {.unnumbered}

This manual serves as a user guide for the **panelView** package in R, which visualizes panel data. For the technical details, see @mou2022.

## Installation

| Source | Version | Date | Features |
|--------|---------|------|----------|
| CRAN | 1.2.1 | 2026-03-20 | Treatment, outcome, bivariate plots |
| GitHub (`master`) | 1.2.1 | 2026-03-20 | Same as CRAN |
| GitHub (`dev`) | 1.3.0 | 2026-03-21 | + `type = "network"`: k-partite graph, singletons, weighted edges |

```{r install-cran, eval = FALSE}
# From CRAN (stable release)
install.packages("panelView")
```

```{r install-github-stable, eval = FALSE, message = FALSE, warning = FALSE, cache = FALSE}
# Stable GitHub version (master branch)
devtools::install_github("xuyiqing/panelView")
```

```{r install-github-dev-branch, eval = FALSE, message = FALSE, warning = FALSE, cache = FALSE}
# Development version (dev branch, includes type = "network")
devtools::install_github("xuyiqing/panelView@dev")
```

```{r check-version}
# Check installed version
installed.packages()["panelView", "Version"]
```

**panelView** depends on the following packages, which should be installed automatically when **panelView** is being installed. You can also install them manually.

```{r install-deps, eval = FALSE}
install_all <- function(packages) {
  installed_pkgs <- installed.packages()[, "Package"]
  for (pkg in packages) {
    if (!pkg %in% installed_pkgs) {
      install.packages(pkg)
    }
  }
}
packages <- c("ggplot2", "gridExtra", "grid", "dplyr")
install_all(packages)
```

The `type = "network"` plot additionally requires **igraph**, which is an optional (suggested) dependency:

```{r install-igraph, eval = FALSE}
install.packages("igraph")
```


## Quick-reference

**panelView** has one main function, `panelview()`, with four core capabilities:

1. **Treatment status & missingness** (`type = "treat"`, `type = "missing"`) — plot which units are treated in which periods, highlight missing values, and summarize treatment histories.

2. **Outcome trajectories** (`type = "outcome"`) — plot the raw outcome variable as time series, colored by treatment status.

3. **Bivariate relationships** (`type = "bivariate"`) — visualize the relationship between a treatment variable and an outcome variable, either in aggregate or unit by unit.

4. **Network structure** (`type = "network"`) — visualize the connectivity structure of panel data as a $k$-partite graph, identifying singletons, connected components, and duplicate observations. Supports two or more sets of fixed effects. See @correia2016 for the graph-theoretic framework.




| Argument | Type | Description |
|---|---|---|
| `formula` | formula | `Y ~ D + X1 + X2`; first RHS variable is the treatment |
| `data` | data.frame | Long-format panel data |
| `index` | character | `c("unit_var", "time_var")`; length $\geq 2$ for `type = "network"` |
| `type` | character | `"treat"` / `"missing"` / `"outcome"` / `"bivariate"` / `"network"` |
| `Y`, `D` | character | Alternative to formula: variable names as strings |
| `by.timing` | logical | Sort units by timing of first treatment |
| `pre.post` | logical | Distinguish pre- and post-treatment periods (binary D only) |
| `collapse.history` | logical | Collapse units with identical treatment histories |
| `show.id` | integer | Plot only units ranked `show.id` (by alphabetical order) |
| `id` | vector | Plot only units whose ID matches `id` |
| `by.group` | logical | Split outcome plot by treatment-status group |
| `by.cohort` | logical | Collapse outcome by treatment-cohort (type="outcome" only) |
| `by.unit` | logical | Plot Y and D per unit (type="bivariate" only) |
| `outcome.type` | character | `"continuous"` (default) or `"discrete"` |
| `leave.gap` | logical | Show time gaps as white bars |
| `axis.lab.angle` | numeric | Angle of x-axis labels: 0, 45, or 90 |
| `legend.labs` | character | Override legend labels |
| `color` | character | Override fill/line colors |
| `legendOff` | logical | Remove the legend |
| `gridOff` | logical | Remove grid lines (continuous treatment) |

Note that *Y*, *D*, and *X* are merely labels; they can be any variables in a panel dataset.

## Shipped datasets

**panelView** ships three datasets for examples and testing:

| Dataset | Description | Units | Periods | Balance |
|---------|-------------|-------|---------|---------|
| `turnout` | US state voter turnout and election reform policies | 47 states | 24 elections (1920--2000) | Balanced |
| `capacity` | State institutional capacity across countries | 47 countries | 44 years (1960--2003) | Balanced |
| `simdata` | Simulated panel with staggered binary treatment | 200 units | 15 periods | Balanced |

All three are balanced panels. The [network structure chapter](#sec-network) uses inline simulation code to construct unbalanced panels that showcase singletons, connected components, and multi-way fixed effects.

## Organization

- [Chapter @sec-treat] — Plotting treatment status and missingness
- [Chapter @sec-outcome] — Plotting outcome trajectories
- [Chapter @sec-bivariate] — Plotting bivariate relationships
- [Chapter @sec-network] — Visualizing network structure: singletons, connected components, and multi-way fixed effects

## Contributors

- [Yiqing Xu](https://yiqingxu.org/){target="_blank"}
- [Licheng Liu](https://liulch.github.io/){target="_blank"}
- [Hongyu Mou](https://hongyumou.github.io/){target="_blank"}
- StatsClaw (Agentic System for Statistical Software Development)

## Report bugs

Please report bugs to yiqingxu [at] stanford.edu or open an issue on [GitHub](https://github.com/xuyiqing/panelview/issues).

<!-- badges: start -->

[![Lifecycle: stable](https://lifecycle.r-lib.org/articles/figures/lifecycle-stable.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [<img src="https://www.r-pkg.org/badges/version/panelView" alt="CRAN status"/>](https://CRAN.R-project.org/package=panelView) [<img src="https://cranlogs.r-pkg.org/badges/grand-total/panelView" alt="downloads: CRAN"/>](https://cran.r-project.org/package=panelView) 

<!-- badges: end -->
