---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# panelview

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**Authors:** Hongyu Mou (PKU); Licheng Liu (MIT); [Yiqing Xu](https://yiqingxu.org/) (Stanford)

**Date:** Jan 28, 2022

**Repos:** [Github](https://github.com/xuyiqing/panelview) (1.1.8)
[CRAN](https://cran.r-project.org/web/packages/panelView/index.html) (1.1.5)

**Examples:** R code used in the tutorial can be downloaded from [here](https://yiqingxu.org/packages/panelview/panelview_examples.R).

---

## Description

**panelview** visualizes panel data. It has three main functionalities:

1. it plots treatment status and missing values in a panel dataset;
2. it plots an outcome variable (or any variable) in a time-series fashion;
3. it visualizes bivariate relationships of two variables by unit or in aggregate.

## Installation

You can install the up-to-date development version from GitHub: 

```{r eval=FALSE}
# if not already installed
install.packages('devtools', repos = 'http://cran.us.r-project.org') 

# currently, v.1.1.8 ("v" is not capitalized)
devtools::install_github('xuyiqing/panelview') 
```

You can also install the **panelview** package from CRAN: 

```{r eval=FALSE}
# currently, v.1.1.5 ("V" is capitalized; we'll update it soon)
install.packages('panelView') 
```

## Tutorial

See [panelview tutorial](https://yiqingxu.org/packages/panelview/articles/panelview.html). For example, plot treatment status in a panel dataset:
```{r example, eval=FALSE}
library(panelview)
panelview(turnout ~ policy_edr + policy_mail_in + policy_motor, 
          data = turnout, index = c("abb","year"), 
          xlab = "Year", ylab = "State")
```

## Issues

Please report bugs to **yiqingxu [at] stanford.edu** with your sample code and data file. Much appreciated!


